---
title: 'Texas Power'
author: "Liane Chen"
date: "2022-10-26"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
“In February 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10–11, 13–17, and 15–20.”[^wiki] For more background, check out these [engineering](https://www.youtube.com/watch?v=08mwXICY4JM&ab_channel=PracticalEngineering) and [political](https://www.youtube.com/watch?v=Zcrsgdl_hP0&ab_channel=Vox) perspectives. 

For this analysis, we will:  
-   estimating the number of homes in Houston that lost power as a result of the first two storms  
-   investigating if socioeconomic factors are predictors of communities recovery from a power outage  

Our analysis will be based on remotely-sensed night lights data, acquired from the [Visible Infrared Imaging Radiometer Suite (VIIRS)](https://en.wikipedia.org/wiki/Visible_Infrared_Imaging_Radiometer_Suite) onboard the Suomi satellite. In particular, you will use the VNP46A1 to detect differences in night lights before and after the storm to identify areas that lost electric power.

To determine the number of homes that lost power, we link (spatially join) these areas with [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84) data on buildings and roads. 

To investigate potential socioeconomic factors that influenced recovery, we will link your analysis with data from the US Census Bureau. 

##### Highlights:
-   load vector/raster data  
-   simple raster operations  
-   simple vector operations  
-   spatial joins  

### Data
#### Night lights
Use NASA's Worldview to explore the data around the day of the storm. There are several days with too much cloud cover to be useful, but 2021-02-07 and 2021-02-16 provide two clear, contrasting images to visualize the extent of the power outage in Texas.

VIIRS data is distributed through NASA's [Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC)](https://ladsweb.modaps.eosdis.nasa.gov/). Many NASA Earth data products are distributed in 10x10 degree tiles in sinusoidal equal-area projection. Tiles are identified by their horizontal and vertical position in the grid. Houston lies on the border of tiles h08v05 and h08v06. We therefore need to download two tiles per date.


#### Roads
Typically highways account for a large portion of the night lights observable from space (see Google's [Earth at Night](https://earth.google.com/web/@27.44405464,-84.7693044,206.63660162a,8916361.52264659d,35y,0h,0t,0r/data=CiQSIhIgMGY3ZTJkYzdlOGExMTFlNjk5MGQ2ZjgxOGQ2OWE2ZTc)). To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas near highways.

[OpenStreetMap (OSM)](https://planet.openstreetmap.org/) is a collaborative project which creates publicly available geographic data of the world. Ingesting this data into a database where it can be subsetted and processed is a large undertaking. Fortunately, third party companies redistribute OSM data. We used [Geofabrik's download sites](https://download.geofabrik.de/) to retrieve a shapefile of all highways in Texas and prepared a Geopackage (`.gpkg` file) containing just the subset of roads that intersect the Houston metropolitan area.\  

- `gis_osm_roads_free_1.gpkg`

[^wiki]:Wikipedia. 2021. “2021 Texas power crisis.” Last modified October 2, 2021. https://en.wikipedia.org/wiki/2021_Texas_power_crisis.

#### Houses
We can also obtain building data from OpenStreetMap. We again downloaded from Geofabrick and prepared a GeoPackage containing only houses in the Houston metropolitan area.\

- `gis_osm_buildings_a_free_1.gpkg`

#### Socioeconomic
We cannot readily get socioeconomic information for every home, so instead we obtained data from the [U.S. Census Bureau's American Community Survey](https://www.census.gov/programs-surveys/acs) for census tracts in 2019. The *folder* `ACS_2019_5YR_TRACT_48.gdb` is an ArcGIS ["file geodatabase"](https://desktop.arcgis.com/en/arcmap/latest/manage-data/administer-file-gdbs/file-geodatabases.htm), a multi-file proprietary format that's roughly analogous to a GeoPackage file.\

We can use `st_layers()` to explore the contents of the geodatabase. Each layer contains a subset of the fields documents in the [ACS metadata](https://www2.census.gov/geo/docs/maps-data/data/tiger/prejoined/ACSMetadata2011.txt).\

The geodatabase contains a layer holding the geometry information, separate from the layers holding the ACS attributes. You have to combine the geometry with the attributes to get a feature layer that `sf` can use. 

## Overview

Below is an outline of the steps we will take to achieve the assignment tasks. 

#### Find locations of blackouts
For improved computational efficiency and easier interoperability with `sf`, we will use the `stars` package for raster handling.\

```{r,  warning = F, message = F,  results='hide'}
# load all packages needed
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(geodata)
library(stars)
library(ggplot2)
library(patchwork)
library(styler)
```


##### combine the data 
-   read in night lights tiles\
-    combine tiles into a single `stars` object for each date (2021-02-07 and 2021-02-16)\
    -   hint: use `st_mosaic`\
    
```{r include=TRUE}
# use read_stars to read in the four night light files
night1 <- read_stars("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif")
night2 <- read_stars("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif")
night3 <- read_stars("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif")
night4 <- read_stars("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif")
```

```{r}
# use st_mosaic to combine the two night light files for each date (create one variable for each date: 2021-02-07 and 2021-02-16)
stars1 <- st_mosaic(night1, night2)
stars2 <- st_mosaic(night3, night4)
```
    
##### create a blackout mask 
-   find the change in night lights intensity (presumably) caused by the storm\
-   reclassify the difference raster, assuming that any location that experienced a drop of more than 200 nW cm^-2^sr^-1^ experienced a blackout\
-   assign `NA` to all locations that experienced a drop of *less* than 200 nW cm^-2^sr^-1^\

```{r include=TRUE}
# find the difference in night lights between the two nights, presumably caused by the storm
night_change <- (stars1 - stars2)

# reclassify the raster and define as 'night_change1". All values less than 200 nW cm^-2^sr^-1^ will be assigned NA by default with the defined breaks
night_change1 <- cut( # use cut() to cut a range of values into bins and specify labels for each bin
  x = night_change, # use difference raster created above
  breaks = c(200, Inf), # define breaks, which auto assigns NA to locations less than 200
  labels = "blackout" # label as "blackout"
) 

# plot to see where there was a change in light intensity between the two nights
plot(night_change1)
```



##### vectorize the mask 
-   use `st_as_sf()` to vectorize the blackout mask\
-   fix any invalid geometries using `st_make_valid`

```{r include=TRUE}
# use 'st_as_sf()' to vectorize the blackout mask and fix any invalid geometries using `st_make_valid`
vect_night_change <- st_as_sf(night_change1) %>% # use the difference raster created above
  st_make_valid(night_change1) # fix any invalid geometries
```


##### crop the vectorized map to our region of interest 
-   define the Houston metropolitan area with the following coordinates\
    -   (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)\
-   turn these coordinates into a polygon using `st_polygon`\
-   convert the polygon into a simple feature collection using `st_sfc()` and assign a CRS\
    -   hint: because we are using this polygon to crop the night lights data it needs the same CRS\
-   crop (spatially subset) the blackout mask to our region of interest\ 
-   re-project the cropped blackout dataset to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)\

```{r include=TRUE}
# define the Houston metropolitan area with coordinates provided in prompt above
houston_coords <- matrix(c(-96.5, 29, -96.5, 30.5, -94.5, 30.5, -94.5, 29, -96.5, 29), ncol = 2, byrow = TRUE)

# create a polygon from the coordinates defined above
houston_p <- st_polygon(list(houston_coords)) %>% # generate a spatial polygon and use coords above
  st_sfc(crs = 4326) # convert the polygon into a simple feature and assign CRS

# crop (spatially subset) the blackout mask to our region of interest
houston_crop <- st_crop(vect_night_change, houston_p) %>% # use houston crop above
  st_transform(3083) # reproject the data to EPSG:3083

```


##### exclude highways from blackout mask 
The roads geopackage includes data on roads other than highways. However, we can avoid reading in data we don't need by taking advantage of `st_read`'s ability to subset using a SQL query.\

-   define SQL query\
-   load just highway data from geopackage using `st_read`\
  -   reproject data to EPSG:3083\
-   identify areas within 200m of all highways using `st_buffer`\
    -   hint: `st_buffer` produces undissolved buffers, use `st_union` to dissolve them\
-   find areas that experienced blackouts that are further than 200m from a highway

`query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"`\
`highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query)`

```{r include=TRUE}
# define SQL query
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"

# load highway data using 'st_read'
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query) %>% # load highway data; use SQL query
  st_transform(3083) # tranform so crs can stack

# identify areas within 200m of all highways using 'st_buffer'
highways_200m <- highways %>% # call in highways data as starting point
  st_buffer(dist = 200) %>% # buffer to specific distance
  st_union() # dissolve buffers

# plot to verify that it works
plot(highways_200m)

# to find the difference with highways data, convert the houston crs
houston_crop_crs <- houston_crop %>% # call in houston cropped data
  st_transform(3083) # convert crs

# find areas that experienced blackouts that are further than 200m from a highway using 'st_difference'
blackout_over_200 <- st_difference(houston_crop_crs, highways_200m)

# check plot to see areas that experienced blackouts that are actually further than 200m from a highway
plot(blackout_over_200)
```


#### Find homes impacted by blackouts

##### load buildings data 
-   load buildings dataset using `st_read` and the following SQL query to select only residential buildings\
  -   hint: reproject data to EPSG:3083\

`SELECT *`\ 
`FROM gis_osm_buildings_a_free_1`\
`WHERE (type IS NULL AND name IS NULL)`\
`OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')`\

```{r include=TRUE}
# define the provided SQL query
query2 <- "SELECT * FROM gis_osm_buildings_a_free_1 WHERE (type IS NULL AND name IS NULL) OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

# use st_read to load the buildings dataset
buildings <- st_read("data/gis_osm_buildings_a_free_1.gpkg", query = query2) %>% # load and select residential
  st_transform(3083) # reproject data to EPSG:3083
```

##### find homes in blackout areas (20 points)
-   filter to homes within blackout areas\
-   count number of impacted homes\

```{r include=TRUE}
# use 'st_filter' to filter to homes within blackout areas
blackout_homes <- st_filter(blackout_over_200, buildings)

# use nrow to count the number of impacted homes
nrow(blackout_homes)
```


#### Investigate socioeconomic factors

##### load ACS data 
-   use `st_read()` to load the geodatabase layers\
-   geometries are stored in the `ACS_2019_5YR_TRACT_48_TEXAS` layer\
-   income data is stored in the `X19_INCOME` layer\
-   select the median income field `B19013e1`\
  -   hint: reproject data to EPSG:3083\


```{r include=TRUE}
# use st_read to load in geodatabase layer for geometries
acs_texas <- st_read(dsn = "data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer = "ACS_2019_5YR_TRACT_48_TEXAS") %>%
  st_transform(3083) # transform the crs

# use st_read to load in geodatabase layer
income <- st_read(dsn = "data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer = "X19_INCOME") %>% # select income data
  select("B19013e1", "GEOID") # select the median income field
```


##### determine which census tracts experienced blackouts 
-   join the income data to the census tract geometries\
  -   hint: make sure to join by geometry ID\
-   spatially join census tract data with buildings determined to be impacted by blackouts\
-   find which census tracts had blackouts\

```{r include=TRUE}
# change column names so they match
income_rename <- income %>% rename("GEOID_Data" = "GEOID") # specify which column to rename and new name

# Join the income data to the census tract geometries by geometry ID
income_census <- left_join(acs_texas, income_rename) # use left join

# verify the class to check in that it is what we want
class(income_census)

# spatially join census tract data with buildings determined to be impacted by blackouts
joined_data <- st_join(blackout_homes, income_census) # use st_join
```


```{r include=TRUE}
# clarify on which census tracts where there were impacts
unique_joined <- unique(joined_data$NAMELSAD) # find unique NAMELSAD values

# check the length
length(unique_joined) # there are 3688 tracts. Exact tracts can be accessed by printing "print(unique_joined) in console

```

##### compare incomes of impacted tracts to unimpacted tracts 
-   create a map of median income by census tract, designating which tracts had blackouts\
-   plot the distribution of income in impacted and unimpacted tracts
-   write approx. 100 words summarizing your results and discussing any limitations to this study

```{r}
# in order to plot the distribution, manipulate income and houston layers and set up unimpacted and impacted variables

# convert crs of income layer
income_census <- income_census %>%
  st_transform(3083) # transform crs

# convert crs of houston layer
houston_p <- houston_p %>%
  st_transform(3083) # transform crs

# create the cropped map that shows houston and income data together
houston_income <- st_crop(income_census, houston_p)
```

```{r}
# create a map of median income by census tract, designating which tracts had blackouts
tm_shape(houston_income) + # call in houston_income data
  tm_fill("B19013e1", title = "Median Income by Census Tract") + # map the median income column, add title
  # tm_layout(inner.margins = c(0.02, 0.02, 0.02, 0.3)) +
  tm_graticules() + # add graticules
  tm_scale_bar(position = c("right", "bottom"), width = 0.2) + # add scale
  tm_compass(position = c("right", "bottom"), size = 0.5) # add compass
```


```{r}
# set up variable for unimpacted tracts in order to plot distribution
unimpacted <- houston_income %>% # use houston income data defined above
  filter(!GEOID_Data %in% joined_data$GEOID_Data) # filter to not include impacted areas

# set up variable for impacted tracts in order to plot distribution
impacted <- houston_income %>% # use the same houston income data
  filter(GEOID_Data %in% joined_data$GEOID_Data) # filter to include impacted areas only

# compare to see that they match
nrow(impacted) == length(unique(joined_data$NAMELSAD))
```

```{r}
# manipulate variables above to prep to show income data

# define the distribution plot 'unimpacted1'
unimpacted1 <- ggplot(unimpacted) + # call 'unimpacted' created above
  geom_histogram(aes(x = B19013e1)) # specify median income to plot

# define the distribution plot 'impacted1'
impacted1 <- ggplot(impacted) + # call in impacted created above
  geom_histogram(aes(x = B19013e1)) # specify income column to plot
```

```{r}
# plot a histogram showing income distribution with both unimpacted and impacted tracts
impacted1 + unimpacted1 # use two variables defined above
```



The plot and histograms show that for affected census tracts, there seems to be more storm impact to those with less income around the $50,000 in annual income mark, with less impact in the higher income range over $200,000 in annual income. This implies that areas in census tracts with higher income levels may be better prepared for intense weather events and less prone to power outages. Their local utility grid may be more elevated or updated and resilient to major storm events. It could be possible that these populations may have more interaction with their utility provider, so that any complaints or needed repairs to electrical equipment are addressed sooner. 


